---
title: "Getting Started"
author: "L.J.M. Aslett, Durham University"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The Mathematics Genealogy Project (MGP), available at <https://mathgenealogy.org/>, is an online platform dedicated to compiling the academic ancestry of mathematicians.
Its mission is to create a comprehensive resource that tracks the intellectual lineage of individuals in the field of mathematics (broadly interpreted) by collecting and organising information about their PhD degrees, advisors, and students.
The project provides access to this genealogical data through their website, enabling researchers and enthusiasts to explore the connections and relationships within the mathematics community.
The MGP plays a vital role in preserving the history and development of the discipline.
We owe a debt of gratitude to Dr Harry Coonce, who founded the project in 1997 ([Jackson, 2007](#refs)), following which the database has [steadily grown](https://www.mathgenealogy.org/growth_image.php) to reach well over 300,000 entries by late 2024.

This package is designed to enable downloading and plotting the genealogical tree data from the MGP in R, expanding on the functionality that is available through the website.

## Getting started

We begin, as always, by loading the package:

```{r setup}
library("maths.genealogy")
```

To utilise the main data-related functions within the package --- specifically, `search_id()` and `get_genealogy()` --- an active internet connection is required.

Every mathematician listed in the Mathematics Genealogy Project (MGP) is assigned a unique ID.
The first task is to locate the ID for the mathematician(s) you wish to investigate.
There are at least two methods to accomplish this. 
The first option is to visit the MGP website's search page at <https://www.mathgenealogy.org/search.php> and complete the search fields.
After locating the desired mathematician and navigating to their data page, you will notice that the URL in your browser's address bar follows the format `https://www.mathgenealogy.org/id.php?id=171971`, where the numbers at the end represent the ID (in this instance, 171971 is the ID for the author of this package).

Alternatively, this package offers a second option that allows you to remain within your R session.
By using the function `search_id()`, you can specify the same search parameters and obtain a data frame summarising the results, including the essential ID for subsequent steps.
Below is the output from a search using the family name of the author of this package:

```{r,eval=FALSE}
search_id("aslett")
#>       id             name                        university year
#> 1 171971    Aslett, Louis           Trinity College, Dublin 2012
#> 2 164179    Haslett, John                                       
#> 3 164818 Haslett, Stephen Victoria University of Wellington 1986
#> 4  57304 Laslett, Geofrey    Australian National University 1975
```

As can be seen, the search performs partial matches and is case insensitive.

If needed, some simple regular expression constructs are also supported.
For instance, if you happened to know the author's family name but only that their first initial is "L", you can narrow down the search by requiring a full match of the family name and a match on the first letter of the given name:

```{r,eval=FALSE}
search_id("^aslett$", "^l")
#>       id          name              university year
#> 1 171971 Aslett, Louis Trinity College, Dublin 2012
```

You will likely be able to find the required mathematician without using regular expressions, but if you would like a beginner's tutorial on how to use them, refer to [Chapter 15 of R for Data Science](https://r4ds.hadley.nz/regexps.html) (Wickham et al., 2023).

## Fetching genealogical data

With the ID `171971` from the search, you can now retrieve the author's genealogical tree:

```{r,eval=FALSE}
g <- get_genealogy(171971)
#> ✔ Connecting to geneagrapher-core WebSocket server [541ms]
#> ✔ Sending query [20ms]
#> ✔ 🎓 Full genealogy retrieved [1.5s]                   
```
```{r,echo=FALSE}
g <- readRDS("171971.rds")
```

When passing only the ID, the function searches the entire tree both "upwards", fetching all ancestors, as well as "downwards", fetching all descendants.
Note that for mathematicians in modern times, there can be a considerable number of ancestors, and for historical mathematicians, the descendant trees may be even larger.
For these reasons, `get_genealogy()` allows you to disable these functions by using the arguments `ancestors` and `descendants` and passing `FALSE`.

## Plotting the genealogical tree

There are currently two ways to plot the genealogical tree.
The first uses Graphviz (Ellson et al., 2004) via the `DiagrammeR` package (Iannone and Roy, 2024), while the second employs the `ggenealogy` package (Rutter et al., 2019).
Neither of these packages is specified as a hard dependency, but you will be prompted to install them if you attempt to use the relevant plotting function without the requisite packages.

Graphviz typically produces the best full-tree view and can be generated using `plot_grviz()`.
Continuing with the previous example, this can be accomplished by:

```{r}
plot_grviz(g)
```

If you are viewing this vignette in a browser, you can zoom in and out using your mouse scroll wheel or trackpad.
Alternatively, you can zoom in by double-clicking and zoom out by holding the shift key while double-clicking.
You can pan around by dragging the plot.

To render this to a PDF file instead of viewing it interactively, simply provide a file name as the second argument:

```{r,eval=FALSE}
plot_grviz(g, "aslett-tree.pdf")
```

As an alternative, `ggenealogy` is excellent for plotting 'local' genealogy, showing a limited number of ancestral/descendant generations, as follows:

```{r,out.width="100%",fig.width=5,fig.dpi=250}
plot_gg(g)
```

You can also control the number of ancestral and descendant generations displayed using `max_anc` and `max_des`:

```{r,out.width="100%",fig.width=7,fig.height=4,fig.dpi=250}
plot_gg(g, max_anc = 11)
```

Care will need to be taken with the sizing of the plot to ensure it is readable: unlike the Graphviz output this relies on R's plot scaling.

## More than one mathematician

Note that the genealogical tree can be built using more than one mathematician's ID when calling `get_genealogy()`.
This will result in each ID being used as a starting point from which to traverse the tree through ancestors and descendants.
The genealogical tree can be constructed using multiple mathematicians' IDs when calling
`get_genealogy()`.
This will use each ID as a starting point for traversing the tree through ancestors and descendants, including all results in a single genealogical tree.

The inspiration for this package arose from a discussion over drinks between the package author and some colleagues at Durham University.
This is as good a way as any to select the next example, which plots the joint genealogical tree built by using those present as starting IDs!

```{r,eval=FALSE}
g <- get_genealogy(c(171971, 108465, 175763, 191788, 169213))
#> ✔ Connecting to geneagrapher-core WebSocket server [558ms]
#> ✔ Sending query [33ms]
#> ✔ 🎓 Full genealogy retrieved [8.5s]
```
```{r,echo=FALSE}
g <- readRDS("171971-108465-175763-191788-169213.rds")
```

Now using the `plot_grviz()` function will plot the *joint* genealogy of the mathematicians specified in the call to `get_genealogy()`.

```{r}
plot_grviz(g)
```

It can be difficult from this plot to identify something of common interest when plotting multiple mathematicians: the nearest common ancestor and shortest path between them.
This can be achieved by using `plot_gg_path()`.
By default, it will plot the shortest path between the first two mathematicians that were specified in the call to `get_genealogy()` which built the genealogical tree, using the x-axis to indicate the year of graduation.

```{r,out.width="100%",fig.width=6,fig.height=3,fig.dpi=250}
plot_gg_path(g)
```

To plot between different mathematicians, one can use the `id1` and `id2` arguments.

```{r,out.width="100%",fig.width=6,fig.height=3,fig.dpi=250}
plot_gg_path(g, id1 = 175763, id2 = 191788)
```

Each id defaults to the first and second mathematicians, so if only one is specified, it overrides that entry only.

```{r,out.width="100%",fig.width=6,fig.height=3,fig.dpi=250}
plot_gg_path(g, id1 = 169213)
```

The above plot includes a common ancestor with a particularly long name (`r "Weierstra\U00DF"`).
Issues like this can be resolved by increasing the value of the `expand` argument above the default, which is 0.15.

```{r,out.width="100%",fig.width=6,fig.height=3,fig.dpi=250}
plot_gg_path(g, id1 = 169213, expand = 0.25)
```


## Other software

Development of this software was inspired by the `geneagrapher` Python package (<https://github.com/davidalber/geneagrapher>) and makes use of the WebSocket server running `geneagrapher-core` (<https://github.com/davidalber/geneagrapher-core>), which was designed to cache queries to the MGP to reduce the load on their main site.
Permission to query the caching endpoint from this R package was [requested and granted in advance](https://github.com/davidalber/geneagrapher/issues/38).

## References {#refs}

Ellson, J., Gansner, E.R., Koutsofios, E., North, S.C. and Woodhull, G. (2004). “Graphviz and Dynagraph — Static and Dynamic Graph Drawing Tools”. In: Jünger, M., Mutzel, P. (eds) _Graph Drawing Software, Mathematics and Visualization_, 127-148. [10.1007/978-3-642-18638-7_6](https://doi.org/10.1007/978-3-642-18638-7_6).

Iannone, R. and Roy, O. (2024). _DiagrammeR: Graph/Network Visualization_. R package, <https://CRAN.R-project.org/package=DiagrammeR>.

Jackson, A. (2007). “A Labor of Love: The Mathematics Genealogy Project”, _Notices of the AMS_, **54**(8), 1002-1003. <https://www.ams.org/notices/200708/tx070801002p.pdf>

Rutter, L., VanderPlas, S., Cook, D. and Graham, M.A. (2019). “ggenealogy: An R Package for Visualizing Genealogical Data”, _Journal of Statistical Software_, **89**(13), 1-31. [doi:10.18637/jss.v089.i13](https://doi.org/10.18637/jss.v089.i13).

Wickham, H., Çetinkaya-Rundel, M. and Grolemund, G. (2024). _R for Data Science: Import, Tidy, Transform, Visualize, and Model Data_. 2nd Edition. O'Reilly Media. ISBN: 978-1492097402.
